### 2025-04-30 Chapters/Chapter1.tex
komið að grein 2.5

%----------------------------------------------------------------------------------------
Wind gusts are brief increase in wind speed (lasting seconds) as compared to mean wind speed. The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period. The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute intervalf. This thesis uses this definition. This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}. The Navier Stokes Equation (\ref{eqn:navierstokes}) shows that the change of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth (the Coriolis force), and frictional force.\cite{uncertainties_in_numerical_weather_predictions}
\begin{equation}
    \label{eqn:navierstokes}
    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
\end{equation}

Traditionally, numerical weather prediction (NWP) systems are used to forecast and analyze weather patterns\cite{medium_range_3d_weather_forecasting_NN}. These models describe the transition between discretized packages of atmospheric states using partial differential equations based on physical reality. These results are usually published every hour, or at courser time intervals for climate simulations. With increasing computer power and efficiency the trend is to output data more often\cite{GNP_vidtal}.They describe the state over the period and so do not necessarily grasp fluctuations well. These fluctuations would include fluctuations in the wind speed, wind gusts\cite{canNNBeatNWP}.

This thesis looks at how best to predict gust factor based on various factors, using several different data sources, including NWP and observations. Being able to accurately predict the wind gust is important as it is often the peak wind gusts that will cause failures in structures. A problem that will become increasingly prevalent in the near future\cite{nasa_extreme_weather}.

\section{Background}
The history of numerical weather predictions goes all the way back to the 1920's when Lewis Fry Richardson pioneered the field and tried to produce forecasts. The results were flawed due to noise in the calculations. ENIAC was built in 1945, it was a general purpose computer that was used, among other things, to make predictions. These predictions took 24 hours to make and were predicting 24 hours into the future. It was a proof of concept but not usable\cite{TheENIACForecastsARecreation}. In the 1950's, with the advent of computers the first operational forecasts emerged. In September of 1954, Rossby and his Stockholm based team produced the first real-time barotropic forecasts. The next year the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton New Jersey, released their first forecasts. These forecasts were for 36 hours at 400, 700 and 900 mb. The results were inferior to subjective human-based forecasts but showed that such forecasts were feasible and promoted further development in the area \cite{historyNWP}. The field of NWP has taken great strides since then following the development of computer power and efficiency.

In the last decade there has been another transformation in the field of weather prediction driven by artificial intelligence. Interest in AI has come in waves. Some progress is made, then interest dwindles. Interest in AI has been increasing steadily since 2010. Notable work that has driven this wave of interest include increase in computational abilities due to parallel processing in graphical processing units (GPU), convolutional neural networks (CNN), which allowed much faster processing of massive (image) datasets and the availability of large datasets online. It is to be noted that images are grid data with some number of channels. Using CNNs could work on any gridded data where there are some spatial features\cite{canNNBeatNWP}. Since 2018, there has been significant work done in the weather prediction field using AI. In 2018, Dueben and Bauer showed that you can build a NN that can outperform a simple persistence forecast and is competitive with very coarse-resolution atmosphere models of similar complexity for short lead times\cite{dueben2018}. Also in 2018, Scher created a deep convolutional neural network (CNN) to emulate a general circulation model (GCM, a numerical model representing the physical processes), training on the GCM which allows it to emulate the dynamics of the model and maintain stability for much longer than Dueben\cite{scher2018}. These two papers were more proof of concept rather than production ready models to replace NWP. They showed that models based on deep learning might, with further development, compete with standard models in the field.

In the last two years there have been even more developments with the emergence of Large AI Weather forecast Models (LWM). In 2024, Ling et al.\cite{SecondRevolution} tried to standardize the definition of LWM in meteorology and came up with 3 rules that need to be met to count as LWM.

\begin{enumerate}[label = Rule \arabic*:]
    \item Large Parameter Count. The number of parameters can vary wildly but a general range might be from tens of millions to billions of parameters
    \item Large Number of Predictands: predicting on different levels (such as pressure levels or height levels) and offering detailed information on the atmospheric vertical structure and surface conditions
    \item Scalability and downstream applicability. This might crystallize in predicting cyclones. Often, the teams responsible for creating these models try to show their applicability to predict cyclones when not trained specifically on cyclone data (e.g. \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast})\cite{SecondRevolution}. This is done to show the versatility of the models.
\end{enumerate}

Before 2022, LWM had been shown to be able to compete with traditional NWP for some specifc cases as well as making predictions quicker, after training. No model had shown that it could in any way completely replace the traditional systems. In early 2022, Pathak et al.\cite{FourCastNet} presented FourCastNet. FourCastNet uses an Adaptive Fourier Neural Operator model that leverages transfomer architecture rather than the popular convolutional model architechture. FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms for smaller variables. It generates a week-long forecast in less than 2 seconds, orders of magnitude faster than standard physical methods\cite{FourCastNet}. In 2022, machine learning methods were presented that made predictions much faster than traditional NWP, after a one time training (or at least training that wouldn't have to be redone often). These were in some cases performing better than NWP. In 2023, Remi Lam and the GraphCast team at Google introduced GraphCast. This model was able to outperform the industry standard High Resolution Forecast (HRES) produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This model as the name suggests leverages graphing connections rather than traditional grid like data structure. The base data is given in latitude and longitude degrees at a resolution of 0.25 degrees. This means points are closer to each other at the poles. Using the graphing structure is supposed to help with bias incurred as a result of this\cite{GraphCast}.

There has been a lot of progress made over the last 6 years (since 2018) and especially in the last 2 years (since 2022)\cite{SecondRevolution}. The progression from machine learning methods being an interesting idea in the field of numerical weather predictions, to outperforming the standard NWP has been remarkably quick. Two years ago, machine learning methods were able to predict quickly and in some niche cases outperform traditional models. They were not generally competitive with standard weather models. Now they are competitive. It is worth noting that the training of these large models is based on data from traditional large weather models. It will be very interesting to watch what the next few years will have in store for the development of machine learning in weather predictions.

\section{Methodology and related work}
This study looks at data from three sources and an attempt is made to predict the gust factor in a given place in Iceland. It uses reanalysis data, along with elevation data to predict the gust factor. It looks at the data at the point of interest. It does not look at the data as a time series. This thesis aims to improve on the baseline model of always predicting the mean gust factor and show that some structure can be learned from reanalysis data about gusts. To do this a neural network was created. Any significant improvement on a base model, that always guesses the mean gust factor, would indicate that the final model has something to contribute.

In 2004, H. Ágústsson and H. Ólafsson\cite{mean_gust_HA_HO} looked at the variability of gust factor in complex landscapes. They looked at data from automatic weather stations that measure wind at 10 meters above ground. The data that was studied in 2004 comes from the same source as used in this thesis, but limits itself to a smaller section. They only looked at the years 1999-2001. They looked at three factors and how these three parameters effected the gust factor. These were $d_m, D, H$, that is direction of wind blowing off a mountain, distance to the mountain and the height of the mountain above the weather station. Their main results were that the gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain. The study in 2004 looked at the effect of a dominant point upwind. It did not look at the effects of the landscape more broadly. In this study, landscape upwind is looked at.

\subsection{Model architechture}
To be able to capture the patterns in the data a neural network was constructed. A NN architechture was chosen as they are known to be able to capture patterns well in complex data and handle high parameter counts. This comes in handy when training on different types of data. It is also easy to construct different types of neural networks and see how they fit well with parts of the dataset. To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE) as defined in Equation (\ref{eqn:mape}) was used.
\begin{equation}
    \label{eqn:mape}
    \text{MAPE} = \frac{1}{n}\Sigma_{i=1}^n\frac{|y_{predict} - y_{true}|}{y_{predict}}
\end{equation}
This was chosen because the target is the gust factor (the wind gust over the average wind). If the target would have been the wind gust rather than the gust factor then something like mean absolute error might be more appropriate.

\subsection{Model explainability}
Neural networks are often considered as mysterious black boxes\cite{nn_black_box}. In an attempt to understand the model predictions, methods designed for explainability are used. One such method is Shapley values\cite{shapley_information}. Shapley values are calculated as the average marginal contribution of a feature value across all possible coalitions. For any combination of parameters what is the contribution of a given parameter. This means that Shapley values can explain individual predictions. Other machine learning tools, like ELI5 (Explain like I am 5), randomly shuffle a feature and look at the effect on model performance\cite{eli5_information}.

>>>>

%----------------------------------------------------------------------------------------
Wind gusts are brief increases in wind speed (lasting seconds) compared to the mean wind speed over intervals of one or several minutes. The gust factor is defined as the peak gust divided by the mean wind speed over a specified time period. The peak wind gust is often defined as the highest 3-second rolling-average wind speed measured over a 10-minute period, while the mean wind speed is the average of all measurements in the same interval. This thesis uses that definition. However, definitions vary: for example, the US uses a 1-minute interval, leading to approximately 14\% higher values \cite{why_wind_gusts}.

The Navier–Stokes equation (\ref{eqn:navierstokes}) shows that changes in wind, both in time and space, depend on the pressure gradient, the oscillating force of the Earth (the Coriolis force), and frictional forces \cite{uncertainties_in_numerical_weather_predictions}.
\begin{equation}
    \label{eqn:navierstokes}
    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{\text{pressure}} -\overbrace{ f\mathbf{k}\times\mathbf{V}}^{\text{Coriolis}} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{\text{resistance}}
\end{equation}

Traditionally, numerical weather prediction (NWP) systems are used to forecast and analyze weather patterns \cite{medium_range_3d_weather_forecasting_NN}. These models describe the evolution of discretized atmospheric states using partial differential equations grounded in physics. Forecasts are typically produced every hour, or at coarser intervals for climate simulations. With increasing computational power and efficiency, the trend is to output data more frequently \cite{GNP_vidtal}. However, such outputs summarize conditions over each period and may not capture short-term fluctuations well, including variations in wind speed and gusts \cite{canNNBeatNWP}.

This thesis examines methods to predict the gust factor using various predictors and multiple data sources, including NWP outputs and observational data. Accurate gust predictions are crucial, as it is ofteh the peak wind gust that will cause damaging incidents, such as structural failures, traffic accidents and downed power lines. Moreover, the prevalence of extreme wind events is expected to increase in the future \cite{nasa_extreme_weather}.

\section{Background}

\subsection*{Numerical weather prediction}
The history of numerical weather prediction dates back to the 1920s, when Lewis Fry Richardson pioneered the field and attempted to produce forecasts. His results were flawed due to numerical noise. The ENIAC, built in 1945, was a general-purpose computer used—among other tasks—for weather prediction. These forecasts took 24 hours to compute and predicted 24 hours into the future. While a proof of concept, they were not operationally useful \cite{TheENIACForecastsARecreation}.

With the advent of electronic computers in the 1950s, the first operational forecasts emerged. In September 1954, Carl-Gustaf Rossby and his Stockholm-based team produced the first real-time barotropic forecasts. The following year, the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton, New Jersey, released its first 36-hour forecasts at 400, 700, and 900 mb. Although these forecasts were inferior to subjective human analyses, they demonstrated feasibility and spurred further development in the field \cite{historyNWP}. Since then, NWP has made tremendous strides in parallel with increases in computational power and efficiency.

\subsection*{AI in weather forecasts}
In the last decade, there has been another transformation in weather prediction driven by artificial intelligence (AI). Interest in AI has come in waves: progress is made, then interest diminishes, but over the past 15 years growth has been steady. Notable drivers of this wave include advances in computational power (notably parallel processing on graphics processing units, GPUs), the developement of neural networks (NNs) for processing massive datasets, and the availability of large online datasets. One type of NN is the convolutional neural network (CNN), originally designed for image processing but applicable to any gridded data with spatial structure \cite{canNNBeatNWP}. Since 2018, significant work has applied AI to weather prediction.

There are two common approaches to AI-based forecasting: combining NWP outputs with NNs, or using NNs alone. In the former, NWP forecasts can be used as inputs to NN training or integrated in other ways; in the latter, NNs are trained directly on meteorological observations, bypassing NWP entirely.

In 2018, Düben and Bauer showed that an NN could outperform a simple persistence forecast and compete with very coarse-resolution atmospheric models of similar complexity for short lead times \cite{dueben2018}. Also in 2018, Scher developed a deep CNN to emulate a general circulation model (GCM)—a numerical model representing physical processes—by training on GCM output. This approach allowed stable emulation of model dynamics for much longer horizons \cite{scher2018}. These papers were proofs of concept rather than production-ready replacements for NWP, but they demonstrated that deep-learning–based models could, with further development, compete with standard models in the field.

\subsection*{Large weather models}
In the last two years, there have been even more developments with the emergence of Large AI Weather forecast Models (LWMs). In 2024, Ling et al.\ \cite{SecondRevolution} proposed a standardized definition of LWMs in meteorology, outlining three criteria, referred to as the "Three Large Rules":

\begin{enumerate}[label=\textbf{\arabic*}.,rightmargin=1.5em]
  \item \textbf{Large parameter count:} Typically ranging from tens of millions to billions of parameters.
  \item \textbf{Multiple predictands:} Forecasting at various levels (e.g., pressure or height levels) to provide detailed atmospheric vertical structure and surface conditions.
  \item \textbf{Scalability and downstream applicability:} Demonstrated, for example, by predicting cyclones even when not explicitly trained on cyclone data (e.g., \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast}) to showcase model versatility \cite{SecondRevolution}.
\end{enumerate}

Before 2022, LWMs had been shown to compete with traditional NWP in specific cases and to generate forecasts much faster after training, No model had been shown to be able to completely replace tradional NWP systems. In early 2022, Pathak et al.\ \cite{FourCastNet} introduced FourCastNet, which employs an Adaptive Fourier Neural Operator model leveraging transformer architecture instead of convolution. FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms them for smaller-scale features. It generates a one-week forecast in under two seconds—orders of magnitude faster than conventional physical methods \cite{FourCastNet}. In 2022, several ML-based models demonstrated faster-than-NWP predictions after one-time training, with performance rivaling or exceeding NWP in some cases.

In 2023, Remi Lam and the GraphCast team at Google released GraphCast, which outperformed the European Centre for Medium-Range Weather Forecasts’ (ECMWF) industry-standard High-Resolution Forecast (HRES). GraphCast uses a graph-based representation instead of a regular grid, operating on a global latitude–longitude grid at 0.25° resolution, which introduces nonuniform point spacing near the poles; the graph structure helps mitigate this bias \cite{GraphCast}.

\subsection*{Conclusion}
Substantial progress has been made since 2018 and especially since 2022 \cite{SecondRevolution}. The evolution from proof-of-concept ML methods to models competitive with—or surpassing—standard NWP has been remarkably rapid. It is important to note that training data for these large models are drawn from traditional NWP outputs, underscoring the synergy between ML and physics-based approaches. The coming years promise further exciting developments in machine-learning–based weather prediction.

\section{Methodology and related work}
In this study, data from three sources is used:

\begin{enumerate}[label=\textbf{\arabic*}.,rightmargin=1.5em]
  \item Three-hourly reanalysis data for Iceland (2004–2023)
  \item Dense (20 by 20~m) gridded elevation data for Iceland
  \item Three-hourly observations from Icelandic meteorological stations
\end{enumerate}

The aim of the study is to improve prediction of gust factors that can be applied at any given place in Iceland. Currently, gust factor predictions can be obtained directly from NWP output, but one may also create a simple regression model (involving, e.g., wind speed and location) to predict them. To accomplish this, an NN is used as a backend to the NWP. The network is trained with gust factors from observations as ground truth, and model quality is measured by comparing predictions with either the NWP or the regression values. Note that only data at the prediction time is used; the time-series aspect of the data is not utilized.

In 2004, Hálfdán Ágústsson and Haraldur Ólafsson \cite{mean_gust_HA_HO} investigated gust factor variability in complex landscapes. They used data from automatic weather stations measuring wind at 10~m above ground during 1999–2001. They examined how three parameters affected the gust factor: \(d_m\), \(D\), and \(H\)—the wind direction relative to a mountain, the distance to the mountain, and the height of the mountain above the station. Their main results showed that the gust factor is inversely correlated with distance and directly correlated with mountain height. Ágústsson and Ólafsson considered the effect of a dominant point upwind but did not consider the broader landscape.

\subsection{Neural networks}
To capture patterns in the data, a neural network was constructed. An NN architecture was chosen because NNs handle complex data well and can manage large numbers of parameters. This flexibility is useful when training on different types of data and allows testing various network designs. An NN uses many matrix calculations to weight input parameters and predict an output. A deep neural network (DNN) has an input layer, an output layer, and one or more hidden layers. Hidden layers transform data from the input layer until it reaches the output layer. The input layer width equals the number of input parameters, and the output layer width equals the number of predictands. Models were created with different parameter sets to gauge each parameter’s influence. Each model produces a single numerical output. Every layer also has an activation function; common choices include Rectified Linear Unit (ReLU), Exponential Linear Unit (ELU), and hyperbolic tangent (tanh), defined in Equations (\ref{eqn:relu}), (\ref{eqn:elu}), and (\ref{eqn:tanh}).

\begin{equation}
    \label{eqn:relu}
    r(x):=\max(0, x)
\end{equation}

\begin{align}
    \label{eqn:elu}
    f(x) := x,\ x > 0\\
    f(x) := \alpha (e^x-1),\ x\leq 0,\ \alpha > 0
\end{align}

\begin{equation}
    \label{eqn:tanh}
    \operatorname{tanh}(x):=\frac{e^x-e^{-x}}{e^x+e^{-x}}
\end{equation}

These activation control neural activation and can help stabilize the network.
\subsection{Model evaluation}
To measure the performance of these models, both in training and testing, mean absolute percentage error (MAPE) as defined in Equation (\ref{eqn:mape}) was used.
\begin{equation}
    \label{eqn:mape}
    \text{MAPE} = \frac{1}{n}\sum_{i=1}^n\frac{|y_{\mathrm{predict}} - y_{\mathrm{true}}|}{y_{\mathrm{predict}}}
\end{equation}
This measure was chosen because the target is the gust factor (the wind gust over the average wind). If the target had been the wind gust rather than the gust factor, mean absolute error might be more appropriate.

\subsection{Model explainability}
Neural networks are often considered mysterious black boxes \cite{nn_black_box}. To understand model predictions, explainability methods are used. One such method is Shapley values \cite{shapley_information}. Shapley values are calculated as the average marginal contribution of a feature value across all possible coalitions. The contribution for a single feature \(j\) to a prediction \(\hat{f}(X)\) is given by Equation (\ref{eqn:shapley}). In Equation (\ref{eqn:shapley}), \(x_j\) is the feature value, \(\beta_j\) its weight, and \(\beta_j E[X_j]\) the mean effect estimate for feature \(j\).
\begin{equation}
    \label{eqn:shapley}
    \phi_j(\hat{f}) = \beta_j x_j - \beta_j E[X_j]
\end{equation}
For any combination of parameters, Shapley values explain the individual prediction by attributing a contribution to each feature. Other methods, like ELI5 (Explain Like I’m 5), randomly shuffle a feature and measure the effect on model performance \cite{eli5_information}.

### 2025-04-30 Chapters/Chapter2.tex
komið að grein 2.5

\chapter{Data gathering and processing}
\label{Chapter2}
Data was sourced from several streams. The Icelandic Meteorological Office (IMO) provided measurements from weather stations all around Iceland. NWP data was downloaded from Copernicus Arctic Regional Reanalysis dataset (CARRA). A land elevation model was also provided by IMO.

@@

IMO provided 10 minute measurements from \nStationsMin weather stations all around Iceland. The measurements that met the filtering criteria, started in \startDateVedur and ended in 2023. Of these \nStationsMin stations, \nVedurMin were from IMO and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 6-7 meters above ground\cite{vegagerdin_postur}. The location of these weather stations can be seen in Figure \ref{fig:aws_map}. The information that is provided by these Automatic Weather Stations (AWS) is presented in two different types of data files, hourly and 10 minute files. The hourly files are summations of the 10 minute files, with the exception that errors, such as nails, still in the 10 minute files should have been removed from the hourly documents. Nails, are sharp increases from the rest of the data and are unrealistic outliers that are considered measurement errors and are discarded. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed ($f$), the wind gust ($f_g$), the standard deviation of the wind gust, the direction of the wind ($d$) and the standard deviation for the wind direction. These measurement started at the end of the 20th century, when the first AWS stations was installed. More have been added in the following decades. In this thesis the data is not considered as a time series, instead the predictions are made using only data at a given time.

\begin{figure}
    \centering
    \includegraphics[scale = 1]{Figures/stationsOverIceland_2024-05-16_stripped_to_frame.png}
    \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study. Most of these were from IMO but over a hundred were from IRCA. IMO AWS are placed at 10 meters above ground, while IRCA stations are placed at around 6-7 meters above ground.}
    \label{fig:aws_map}
\end{figure}

\section{CARRA Data}
The CARRA dataset goes back to September 1991 and is currently updated monthly, with a latency of 2-3 months\cite{carra_information}. The oldest IMO data point that fulfills given criteria of consistent available datapoints is from \startDateVedur. This is covered by CARRA. The CARRA dataset is available for two regions, West and East. These regions cover a large part of the North Atlantic, including Greenland and Scandinavia. Each of these covers a vastly larger area than the area of interest. This leads to having to store a large amount of data. To get the data one has two options. CARRA web interface or using the API client provided by CARRA. Using the API client is the only realistic option here, as there were thousands of requests made for different times. If using the API, it is possible to query a smaller area (such as a rectangular area around Iceland) given a set of coordinates.

The requests to the API are made at each available CARRA hour ([00, 03, 06, 09, 12, 15, 18, 21]) for each available observation. Only these hours can be used as the CARRA predicted output is represents the wind speed at those given hours and would thus not encapsulate well the extreme values in between these times. Using these datapoints, interpolation was used to get an estimation for the point of the given weather station. The CARRA data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this thesis was downloaded from height levels. That is, data was requested at heights of 15, 250 and 500 meters above ground. For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be interpolated to create data for model to be trained on, as the CARRA grid points do not coincide with weather stations.

\section{Elevation data}
IMO provided a TIFF file containing the elevation of Iceland on a 20 meter by 20 meter grid. This file encompasses Iceland and is around 685 MB. The Python package Rasterio allowed for quick lookup with it's index and the affine transform. Using this package it is possible to quickly look up elevation given coordinates using matrix calculations. This package allowed for lookup by coordinates and by positioning inside the grid encompassing Iceland. It also allows for index lookup using coordinates. Using index lookup, neighboring points can be looked up and used to bridge the elevation for exact posisiton of given point.

\label{Chapter3} 
\section{Combining data sources}

This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models. When working with hundred of thousands of rows, the efficiency of the code is very important. Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible. The three data sources were all in a different format. Measurement data from IMO was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format. To use the data to train, these three data sources needed to be combined into one file. This was done based on the measurement data from the IMO. A limit was set on the average wind speed and it was used to select measurement points. Along with the average wind speed having to be above a certain limit, to ensure that the same weather for the same location is not duplicated. The data from IMO was supplied for 10 minute increments, while CARRA data is in 3 hour intervals. This means that to use the CARRA data to predict the measured values from IMO, temporal interpolation would need to be done. Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5 km while the the information from the IMO is given at specific locations. The elevation information was given by a 20 by 20 rectangular grid that covers Iceland. When combining these data sources an interpolation method needs to be decided upon. Here linear interpolation was applied, both temporally and spatially. As the measurement points chosen were not outliers, that is the strongest average wind speed in a given range, interpolating in time does not work well. This necessitates using only measurements that are exactly at three hour intervals (00, 03, 06, 09, 12, 15, 18, 21 hours) at interpolating only in space. The choice of interpolation method, although potentially impactful on the results, was not specifically addressed in this study.

\begin{figure}[h]
    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
    \caption{A flow chart showing how data sources were combined}
    \label{fig:data_preprocessing_flow_chart}
\end{figure}

The procedure of combining these sources was as follows and can be seen in Figure \ref{fig:data_preprocessing_flow_chart}. The measured data from the AWS is filtered by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed due to higher wind gusts. After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API. The CARRA API needs to be queried for given hours, days, months, years and a given area. That is, if queried for a given hour, it returns that hour for every day that is queried. Similarly if queried for a given day, it returns that day for every month. In light of these restraints, it was decided to query month by month. Querying only the days needed but every hour of the day (UTC 00, 03, 06, 09, 12, 15, 18 and 21). After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe. After this the downloaded data is discarded and the next month is queried. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (a reduction from several terabytes to less than a gigabyte).

Once CARRA data has been merged with AWS data, using station and time columns, then this combined file needs to checked for nails. This is done by using the hourly data (which is supposedly error free). As the data has been filtered in for the highest average wind speed in a 48 hour interval, the hourly data can be used to find nails. The hourly data is combined with the merged AWS and 10 min data. Then filtering is applied on the average wind. If the average wind speed for the 10 minute data is higher than the hourly the rows are dropped. Most stations have nails in less that 10\% measurements. The stations that have higher than 10\% error are ignored.

The elevation data comes in a GeoTIFF file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved interpolation needs to be done. This is done in a similar manner to the interpolation of the CARRA data. The four points bounding the point of interest were used to linearly interpolate the value of the point of interest. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

>>>>

\chapter{Data gathering and processing}
\label{Chapter2}
Data were sourced from several streams. The Icelandic Meteorological Office (IMO) provided measurements from weather stations across Iceland, NWP data were downloaded from the Copernicus Arctic Regional Reanalysis dataset (CARRA), and finally a land-elevation model was also provided by the IMO.

@@

IMO provided measurements from \nStationsMin weather stations across Iceland. The measurements that met the filtering criteria began in \startDateVedur and ended in 2023. Of these \nStationsMin stations, \nVedurMin were from the IMO, with the anemometer at 10~m above ground, while the remaining \nVGMin stations were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)}, with the anemometer at 6–7~m above ground \cite{vegagerdin_postur}. The locations of these weather stations are shown in Figure \ref{fig:aws_map}.

Data from these automatic weather stations (AWS) are stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases. Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed (\(f\)); wind gust (\(f_g\)); standard deviation of the wind gust; wind direction (\(d\)); and standard deviation of the wind direction.

These measurements began at the end of the 20th century with the installation of the first AWSs, and more stations have been added in subsequent decades.

\begin{figure}
    \centering
    \includegraphics[scale = 1]{Figures/stationsOverIceland_2024-05-16_stripped_to_frame.png}
    \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study. Most of these were from IMO but over a hundred were from IRCA. IMO anemometers are placed at 10 meters above ground, while IRCA ones are placed at 6--7 meters above ground.}
    \label{fig:aws_map}
\end{figure}

\section{CARRA Data}
CARRA is a high-resolution atmospheric reanalysis produced by the Copernicus Climate Change Service and run by ECMWF. It covers two regions, a west region covering Greenland and Iceland and an east region covering the European Arctic. It has a 2.5 km horizontal resolution and dates from 1991 to the present, with monthly updates. CARRA provides three-hourly analysis fields and short-term forecasts (hourly for lead times under 6 h and three-hourly beyond) of surface and near-surface variables—wind, temperature, pressure, precipitation, etc. It is based on the HARMONIE-AROME limited-area NWP model, forced at its boundaries by ERA5 (ECMWF Re-Analysis v. 5) and enhanced by local observations to better represent complex terrain, land–sea contrasts, and sea-ice processes. It is updated monthly, with a latency of 2-3 months\cite{carra_information}.

The CARRA dataset covers all the IMO observations that fulfill criteria of consistent availability – the oldest observation is from \startDateVedur. The CARRA-West region covers a vastly larger area than the area of interest. This leads to having to store a large amount of data. To download CARRA data one has two options, a web interface or using an API client provided by CARRA. Using the API client is the only realistic option here, as there are thousands of requests made for different times. If using the API, it is possible to query a smaller area (such as a rectangular area around Iceland) given a set of coordinates, but this is not possible with the web interface.

The requests to the API were made at each available CARRA hour ([00, 03, 06, 09, 12, 15, 18, 21]) on a grid covering Iceland, for each available observation time. The downloaded data were interpolated to get values at the weather stations. CARRA contains several types of layers: single levels, model levels, height levels, and pressure levels. The data for this thesis was downloaded from height levels. They were requested at heights of 15, 250 and 500 meters above ground. For each point 4 parameters were requested, wind speed, wind direction, pressure, and temperature.

\section{Elevation data}
A GeoTIFF file containing Iceland’s elevation on a 20~m by 20~m grid was provided by the Icelandic Meteorological Office (IMO). The entire country is covered by this file, and its size is approximately 685\,MB.

The Python package \texttt{rasterio} is used, enabling rapid elevation lookup via its spatial indexing and affine‐transform capabilities. Elevation at specified geographic coordinates can be retrieved directly, and grid indices may be used for efficient access. Elevation for any exact location can be interpolated by fetching neighboring grid points.

\section{Combining data sources}

Three main data sources were used, each requiring querying, filtering, and merging to prepare the combined dataset. When handling hundreds of thousands of rows, code efficiency is essential: row-by-row iteration can increase execution time dramatically compared to vectorized operations.

The sources were provided in different formats: IMO measurement data in text files, elevation data in GeoTIFF, and CARRA reanalysis data in GRIB. To train the models, these datasets were combined into a single file using the IMO measurements as a reference. CARRA data are supplied on a rectangular grid with approximately 2.5~km spacing, while IMO observations are tied to specific station locations. Elevation data are on a 20~m by 20~m grid covering Iceland. Linear interpolation was applied to merge the sources.

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{Figures/data-preprocessing-flow-chart.png}
  \caption{Flow chart illustrating the data-combination workflow}
  \label{fig:data_preprocessing_flow_chart}
\end{figure}

The merging procedure, shown in Figure \ref{fig:data_preprocessing_flow_chart}, was as follows: for each AWS observation, a query was constructed for the CARRA API by specifying year, month, day, hour, and spatial extent. Because the API returns all specified days when queried by hour, and all specified months when queried by day, monthly queries were issued for only the required days, retrieving all eight three-hourly time points (00, 03, 06, 09, 12, 15, 18, and 21~UTC). After downloading the requested variables at the desired pressure levels, point values were interpolated and appended to a pandas dataframe. The monthly GRIB files were then discarded before proceeding to the next month. This strategy reduced storage needs from several terabytes to under one gigabyte.

Elevation values from the GeoTIFF were interpolated in the same manner: the four surrounding grid points were used in a linear interpolation to estimate the elevation at each station location. These interpolated values were included in the dataframe, since topography influences both average wind speed and gustiness \cite{GNP_vidtal}.

\section{Comparison of observed and reanalysis wind speed}
